---
title: "Here Comes the Sun: The Effect of Earlier Sunrises and Sunsets on Academic Achievement"
author: "Hannah M. Stackpole"
format: revealjs
editor: source
execute: 
  cache: true
---
## Literature TL;DR {.smaller}

- DST moves a location one time zone father east
- Later sunsets result in less sleep and more disease (Giuntella & Mazzona, 2019)
- Later school starts (better for adolescents) result in higher grades, more time on HW, less time on TV, less day-time sleepiness, fewer depressive symptoms (Carrell, 2011; Edwards, 2012; Heissell & Norris, 2018; Wahlstrom, 2002; 2017) 

## Conceptual Model {.smaller}

- Three clocks: sun, body, social (Roenneberg et al., 2019)
- Sunsets & sleep timing: especially important for adolescents
- Sleep & test scores
  - Less sleep (chronic sleep deprivation)
  - Later sunrises (drowsy in mornings)
- Long term implications: human capital, wages, and life outcomes (Becker, 1964; Mincer, 1974; Card, 2000; Hanushek & Woessmann, 2008; 2012; Chetty et al., 2014; Mulligan et al., 1999; Murnane et al., 1995; 2000).
  - educational achievement matters (not just attainment)

## Empirical Framework {.smaller}

:::: {.columns}

::: {.column width="50%"}
```{r}
source("07. analysis.R")
proficiency_EMH_year_plot +
  labs(title = "Subject Proficiency Rates for \nElemenentary, Middle, and High Schoolers")
```
:::

::: {.column width="50%"}

```{r}
source("07. analysis.R")
map_2006 +
  labs(title = "2006 Time Zone Change",
       fill = "County Designation",
       caption = "Treated counties Daviess, Dubois, Knox, Martin, Perry, and Pike move to Central time zone on April 2, 2006 \nwhile surrounding control counties (green) remain in the Eastern time zone")
```
:::

::::

- Economists' favorite: DiD
- ${Y}_{itg} = \beta_1(\text{treated}_{ig} \times \text{post_treat}_t) + \alpha_i + \gamma_t + \delta_g + \varepsilon_{itg}$
- treatment counties move to Central time zone in April 2006


## Main Results {.smaller}
```{r}
main_results_tbl
```

- basically nothing (potential effects in ELA/HS)
- ELA estimates increase as age increases which is consistent with the literature and my hypothesis 

# Heterogeneous Effects {.smaller}
```{r}
prof_level_results_tbl |>
  gt::tab_options(
    table.font.size = px(11),
    data_row.padding = px(2)
  )
```
- I went a bit overboard lol -> p-val corrections :(
- **ELA / All / Very Low** & ELA / Low / V Low promising; consistent with Edwards (2012)


# Conclusion? {.smaller}
- may or may not be anything here
- limited data and ability to establish pre-treatment parallel trends
- would have liked to do more and am still working :/ 


